---
title: "Reproducible Research: Peer Assessment 1"
author: "Maria Freydlin"
output: 
  html_document:
    keep_md: true
    self_contained: no
    fig_caption: yes
    toc: yes
  
---


## Loading and preprocessing the data
```{r load, warning=F, message=FALSE}
library(dplyr)
library(ggplot2)
unzip("activity.zip")
activity <- read.csv('activity.csv')
activity$date <- as.Date(activity$date, '%Y-%m-%d')
str(activity)
```

## What is mean total number of steps taken per day?

Calculate total daily steps
```{r total, warning=FALSE, message=FALSE}
step_sum <- activity %>% group_by(date) %>% summarise(steps_per_day=sum(steps, na.rm=T))
head(step_sum)
```
Plot a histogram of total daily steps
```{r plot} 
g <- ggplot(step_sum, aes(steps_per_day)) +
geom_histogram(fill='blue', color='black', bins=37)+theme_bw()
print(g)
```

Calculate average daily steps

```{r average, message=FALSE}
mean(step_sum$steps_per_day)
```

Calculate Median Daily Steps

```{r median, message=FALSE}
median(step_sum$steps_per_day)
```


## What is the average daily activity pattern?

Calculate average steps in each interval
```{r pattern, message=FALSE}
int_avg_steps <- activity %>% group_by(interval)%>% summarise(avg_steps_int=mean(steps, na.rm=T))
head(int_avg_steps)
```

Time series plot of average steps by interval

```{r plot2}
t <- ggplot(int_avg_steps, aes(x=interval, y=avg_steps_int))+geom_line()+theme_bw()
print(t)
```

Find interval with maximum steps

```{r max}
max_steps_int <- filter(int_avg_steps, avg_steps_int==max(avg_steps_int))
print(max_steps_int)
```


## Imputing missing values

Total number of missing values

```{r missing}
missing_values <-sum(!complete.cases(activity))
print(missing_values)
```

Impute missing steps by the average steps in each interval

Create interval average values

```{r intavg}
int_avg <- function(interval){
         return(mean(activity$step[activity$interval==interval], na.rm=T))
}
int_a <- sapply(activity$interval, int_avg)
head(int_a)
```

Impute steps using int_avg values

```{r impute}
impute_steps <- function(steps, int_a){
        out <- steps
        for (i in 1:length(steps)){
                if (is.na(steps[i])){
                        out[i] <- int_a[i]
                } else {
                        out[i] <- steps[i]
                }
        }
        return(out)
}
steps <- impute_steps(activity$steps, int_a)
head(steps)
```

Create new data set with imputed steps

```{r new}
activity.new <- activity
activity.new$steps <- steps
head(activity.new)
```

Histogram of the new data reflects that missing data was removed and due to the imputation mean and median are now the same and more accurate

```{r newhist, message=F}
step_sum_new <- activity.new %>% group_by(date) %>% summarise(steps_per_day=sum(steps))
head(step_sum_new)
n <- ggplot(step_sum_new, aes(steps_per_day)) +
        geom_histogram(fill='blue', color='black', bins=37)+theme_bw()
print(n)
```

Average of the imputed daily steps 

```{r avgnew, message=F}
mean(step_sum_new$steps_per_day)
```

Median of the imputed daily steps

```{r medianew, message=F}
median(step_sum_new$steps_per_day)
```


## Are there differences in activity patterns between weekdays and weekends?

Create new variable for weekdays and weekends

```{r newvar, message=F}
days <- weekdays(activity.new$date)

wkdays <- function(days){
        out <- days
        for (i in 1:length(days)){
                if (days[i] %in% c('Saturday', 'Sunday')){
                        out[i] <- 'weekend'
                } else {
                        out[i] <- 'weekday'
                }
        }
        return(out)
}
activity.new$day <- wkdays(days)
activity.new$day <- as.factor(activity.new$day)
str(activity.new)
```

Plot average steps by interval on weekdays and weekends

```{r plotw, message=F}
int_avg_steps_new <- activity.new %>% group_by(interval, day)%>% summarise(avg_steps_int=mean(steps, na.rm=T))
head(int_avg_steps_new)

w <- ggplot(int_avg_steps_new, aes(x=interval, y=avg_steps_int))+
        geom_line()+facet_grid(day ~ .)+theme_bw()
print(w)
```

